---
title: "Homework 4: Differential gene expression analysis (T-test) in R"
author: "Aizhan Uteubayeva. NetID:au198"
date: "2023-09-27"
output: html_document
---
## Step 1- Read in data into R. Check if data was imported correctly
### 1a - Read in clinical data
```{r}
library(readr)
c.clinData <- read.table(file = "CRC_PILOT_clinical_data_HIDS.tsv",
                       sep="\t", 
                       header = T, 
                       stringsAsFactors = F, row.names = NULL)
knitr::kable(head(c.clinData[,1:4],n = 6))
```
### 1b: Read in processed normalized gene expression data in log2 scale, includes gene annotation
```{r}
c.geneExp <- read.table(file = "CRC_PILOT_withGeneAnno.tsv",
                       sep="\t", 
                       header = T, 
                       stringsAsFactors = F, row.names = 1)
# transposed the gene expression data
c.geneExp <- as.data.frame(t(c.geneExp))

knitr::kable(c.geneExp[1:5, 1:4], caption = "Processed gene expression data in log2 scale with gene annotation")
```
### The Goal is to compare which two groups ?
Our goal is to compare 20 patients that relapsed/recurred with 20 patients that did not relapse
-   Question: What is the outcome variable column from clinical data file ? It is the column in the clinical file that contains the baseline and comparison values
-   Answer: "RECURRENCE_ANY" column has values "NO" (baseline group) and "YES" (comparison group)
```{r}
knitr::kable(table(c.clinData$RECURRENCE_ANY))
```
## Step 2 - Clean/Filter data

### Filter clinical data - not needed

Clinical data has one row per patient. So cleaned and ready to go

```{r}
dim(c.clinData)
head(c.clinData)
```
### Filter gene exp data

Check rows and columns:

-   Rows are features (genes in this case)
-   Columns are patients
```{r}
knitr::kable(c.geneExp[1:5, 1:4]) 
```
### leave only the tumor samples in gene expression table = 40 samples
```{r}
library(dplyr) 
c.geneExp2<- c.geneExp %>% select(-contains("Normal"))
knitr::kable(c.geneExp2[1:5, 1:4]) 
```
### exract the ID for the patients from gene expression data
```{r}
#Check the col names of your gene expression object
head(colnames(c.geneExp2))

# Function definition - In this function I define my string split conditions based on underscore as a delimiter
funSplit <- function(x1) {
    
    x2 <- unlist(strsplit(x = x1 , split = "_", fixed = TRUE))
    x3 <- x2[2]
    return(x3)
}

# Calling the function using apply(). I am asking the function to apply string split on every sample ID and then return the cleaned sample ids 
newNames1 <- apply(X = as.matrix(colnames(c.geneExp2)), 
                       MARGIN = 1,
                       FUN =  funSplit)

#new cleaned sample IDs
head(newNames1)

# You can set this as the new rownames for your gene expression object
colnames(c.geneExp2) <- newNames1
knitr::kable(c.geneExp2[1:5, 1:4]) 
```
### What are the sample identifiers ?

The GSM ids are the identifiers in the gene exp data that match the identifiers for the Clinical data file:

-   Sample identifier for clinical data: Column called "SUBJECT_ID"
-   Sample identifier for gene expression data: Column names are the IDs

```{r}
matchingSamples <- which(colnames(c.geneExp2) %in% c.clinData$SUBJECT_ID) # 40 IDs matching
subsetGeneExp <- c.geneExp2[, matchingSamples] ## 17881 rows features, 40 patient columns
```
## Step 3 - Identify the groups to be compared

### Identify the groups to be compared (Baseline and Comparison Grps)

```{r}
# Labels (row numbers) that can identify the baseline group patients
baselineGrpLabels <-  which(c.clinData$RECURRENCE_ANY == "NO") 
head(baselineGrpLabels)

# Use the labels (row numbers) to subset baseline patients in clinical data file = 20 patients
c.clinBase <- c.clinData[baselineGrpLabels, ] 

# Labels (row numbers) that can identify the comp group patients = 20 patients
compGrpLabels <- which(c.clinData$RECURRENCE_ANY == "YES") 
head(compGrpLabels)
# Use the labels (row numbers) to subset comp patients in clinical data file 
c.clinComp <- c.clinData[compGrpLabels, ]

#### Use the clinBase and clinComp objects to subset gene expression data
c.geneExpTumorBase <- subsetGeneExp[, c.clinBase$SUBJECT_ID] # 41927 feature (rows), 6 samples columns
c.geneExpTumorComp <- subsetGeneExp[, c.clinComp$SUBJECT_ID] # 41927 feature (rows), 16 samples columns
```
## Step 4: Sanity check

-   See if filtering of clinical data in R matches filtering of clinical data in excel
-   See if sample ids in clinical data match sample ids in gene exp data (if they don't match it means your step 1 and/or 2 is wrong)
-   Verify you see correct number of samples in baseline and comp groups
-   Export the column names from gene expression data to see if it contains only probe/gene names and no other garbage

```{r}
#Looking if sample ids in clinical data match sample ids in gene exp data
c.clinBase$SUBJECT_ID == colnames(c.geneExpTumorBase)
c.clinComp$SUBJECT_ID == colnames(c.geneExpTumorComp)

### Checking if the samples IDs baseline and comp groups are the same 
write.table(c.clinBase$SUBJECT_ID, file = "output/CRC ClinBaseIDs.tsv", sep="\t", quote = F  )
write.table(c.clinComp$SUBJECT_ID , file = "output/CRC ClinCompIDs.tsv", sep="\t", quote = F  )
write.table(colnames(c.geneExpTumorBase) , 
            file = "output/CRC GeneExpBaseIDs.tsv", 
            sep="\t", quote = F  )
write.table(colnames(c.geneExpTumorComp) , 
            file = "output/GeneExpCompIDs.tsv", 
            sep="\t", quote = F  )

#Exported the features from gene expression data  
#Opened this file and check that it contains only probe/gene names and no other garbage
write.table(rownames(c.geneExp),file = "output/CRCgenExpFeatureIDsCheck.tsv", sep="\t", quote = F  )
```
## Step 5: Prep data for T-test
```{r}
### Checking to make sure data is a numeric data frame
knitr::kable(head(c.geneExpTumorBase[1:5,1:4]))
knitr::kable(head(c.geneExpTumorComp[1:5,1:4]))
```
#### Function for T-test
```{r ttest}
source("fnTTest.R")

#### Call T-test function
results1 = fnTTest(baseGroup = c.geneExpTumorBase, 
                   compGroup = c.geneExpTumorComp, 
                   testName = "CRC_TTest_", 
                   baseGroupName = "Non-relapsed",
                   compGroupName =  "Relapsed",
                   folderName = "output")
```
## Brushing up
### creating two csv finles with ordered p-value>0.01 and top 20 genes
```{r}
c.ttest <- read.csv(file = "CRC_TTest.csv")
p = c.ttest[c.ttest$Pvalue < 0.01, ] # 146 genes that have p-value lower than 0.01
p.extracted <- data.frame(c(p$Feature), c(p$Pvalue)) # extracted features and p-value
p_final <- p.extracted[order(p.extracted$c.p.Pvalue.),] # ordered by ascending value of p
colnames(p_final) <- c("Features","p-value") # changed the name of the columns

# saving the df into csv file
write_csv(p_final,"/Users/utb_aizhan/Desktop/R projects/HIDS7003_Precision_Health/GroupComp_Rproj/GroupComp/p_final.csv" )

# selecting only 20 top genes
final_features <- head(p_final,20)
write_csv(final_features,"/Users/utb_aizhan/Desktop/R projects/HIDS7003_Precision_Health/GroupComp_Rproj/GroupComp/final_features.csv" )
```


